Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-23 draft.tex
.

%\makeatother
%\MSonehalfspacing

@@

\begin{document}
\maketitle

\section{Introduction}
Among many functional properties of the mammalian central nervous system, it is involved in maintaining many physiological properties in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found up till now, and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though, experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons being more or less active than a certain set point of mean activity are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. Furthermore, one can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing "intrinsic" properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}. It is important to note that - depending on the specific mechanisms involved - both concepts of neural homeostasis can occur on a very wide range of different time scales, ranging from milliseconds to hours or days.

@@

Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity which has previously been suppressed by the rigid single-cell homeostasis. In the face of possible \emph{new} features within the network's structure, one should further clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

Equation \eqref{NO_dyn} describes the influx of NO as a sum of scaled and spatially shifted Dirac functions. Apart from the question, whether this source term results in a well defined, finite solution at the neurons' positions (see section !!??!!), it can only be modeled to a certain degree of accuracy, depending on the resolution of the numeric grid. In practice, we approximated the point sources of NO as an insertions at individual grid cells at a rate of $nNOS^i(t) / h^2$, where the normalizing divisor $h^2$ ensured the desired total influx per neuron. This numeric implementation required two additional constraints: First, all random neuron positions were confined to integer multiples of $h$ in x- and y-direction. Second, to avoid redundancy and for physiological reasons, each grid cell could only hold one neuron at maximum. Both conditions combined led to an iterative generation of positions, where for each neuron a random number generator produced pairs of integers $(n_x,n_y)$, each within $[0,N]$, until an unoccupied pair of integers was found and occupied, moving on to the next neuron. Figure \ref{diff_test_plot} shows an example of the resulting NO density. 
\begin{figure}[h!]
%% density_plot.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/density_plots/density_plot.png}
\end{center}
\caption{Example of NO-diffusion with 400 point sources of excitatory neurons.}
\label{diff_test_plot}
\end{figure}

@@

\subsection{Activity Analysis} \label{activ_analys}
As a first attempt, we set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after 750 seconds, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group, the average $NO$ concentration at the excitatory neurons' positions and the average firing threshold within the excitatory group (inhibitory threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in the upper three plots of Figure \ref{full_sim_osci}, are in fact regular oscillations across all three depicted variables. While the oscillation amplitudes undergo a rather unpredictable time course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. This feature is also illustrated by the fact that the power spectra depicted in Figure \ref{Power_Spec_without_Analysis} have a peak at the frequency that appeared to be dominant in Figure \ref{full_sim_osci}. The fact that the mean over all power spectra of excitatory thresholds differs from the power spectrum of the mean of these thresholds in its amplitude indicates that not all thresholds oscillate at the same phase. Still though, the overall shape of both spectra are equivalent. 
\begin{figure}

\begin{center}
%% oscillating_activity_plot.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
\end{center}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at excitatory neurons' positions. C/C*: Average excitatory firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}
\begin{figure}
\begin{center}
%%% plot_pow_spec.py
\includegraphics[width=0.7\textwidth]{../../plots/power_spec/power_spectrum_sim_only.png}
\caption{Blue: Mean of power spectrum over all excitatory thresholds. Red: Power spectrum of mean over all excitatory thresholds.}
\label{Power_Spec_without_Analysis}
\end{center}
\end{figure}
Although one might argue that - for the sake of practical purposes - regular oscillations of the $NO$ concentration are not of much concern for the network as such, a periodically fluctuating firing rate is a qualitative feature requiring further inquiry. Rhythmic patterns of activity are a commonly observed phenomenon within the brain. Depending on the rate of oscillations, the location within the brain and the spatial scale over which they are observed, different physiological features are found to be responsible, leading to explanatory models at different levels of abstraction. On a microscopic, single-neuron level, a so-called slow after-hyper-polarization conductance explains the appearance of bursting behavior: A relatively short period of spiking activity is followed by a longer period of silence. This phenomenon can be modeled by a $Ca^{2+}$-dependent $K^{+}$-conductance. Upon the bursting period, $Ca^{2+}$ concentration rises, prohibiting continuous spiking due to the activation of $K^{+}$  current until the $Ca^{2+}$ concentration has returned to base-level \cite[p.~203-207]{Theor_Neur_Dayan}. However, due to the simplicity of our neuron model, this type of oscillating activity can be ruled out as an explanation.

On a larger, rate-based level, dynamical models of interconnected excitatory and inhibitory groups can exhibit sustained oscillations of activity \cite[p.~270]{Theor_Neur_Dayan}. As a simplest model, one can describe the dynamics of the excitatory and inhibitory populations as a 2-dimensional dynamical system, representing the respective population rates $r_e$ and $r_i$:
\begin{align}
\tau_e \dot{r}_e  &= -r_e + \phi_e(r_e W_{ee} + r_i W_{ie}) \label{2d-exc_inh_rate_1} \\
\tau_i \dot{r}_i  &= -r_i + \phi_i(r_e W_{ei} + r_i W_{ii}) \label{2d-exc_inh_rate_2}
\end{align}

where $\tau_e$ and $\tau_i$ represents time constants representing the rapidity of firing rate adaptation, $\phi_e$ and $\phi_i$ the respective transition functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the 2-dimensional phase plane. According to the Poincaré-Bendixson theorem, such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing "into" $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.
Coming back to the dynamics in question, one would need to show that the network has a locally unstable fixed point at $\mathrm{(r_e,r_i)\approx (3\, Hz, 13\, Hz)}$, see A* in Figure \ref{full_sim_osci}. Several aspects come into play, making it a hard task to fit the full spiking network model onto the simplified equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. First, the choice of $\tau_e$ and $\tau_i$ in \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} has significant influence onto the stability of the system. However, a straightforward equivalent does not exist in the used spiking neuron model. It has been shown that in the case of slowly varying input $\tau_{e/i}$ of equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} are equal to the membrane constant $\tau_m$ of the LIF-neuron used in the network \cite{Gerstner_Pop_Act}, see equation \ref{LIF_Dynamics}. A second problem is to choose a good estimate of the activation functions $\phi_{e/i}( \cdot )$. In the absence of noise in equation \ref{LIF_Dynamics}, the firing rate for a given constant total input $J$ is
\begin{equation}
\phi(J) = \begin{cases}\frac{1}{-\tau_m ln(\frac{V_t -E_l-J\tau_m}{V_r -E_l-J\tau_m})} & \mathrm{for}\; J\tau_m + E_l > V_t\\
0 & \mathrm{otherwise}
\end{cases}.
\label{LIF_Fir_Rate_no_Noise}
\end{equation}
If noise $\sigma$ is present, according to \cite{Roxin_Firing_Rate_Distribution} one can calculate the \emph{mean} firing rate by
\begin{align}
\phi(J) &= \left[ \sqrt{\pi}\tau_m \int_{x_-}^{x_+} dx e^{x^2} erfc(-x) \right]^{-1} \label{LIF_Fir_Rate_with_Noise1}\\
x_- &= (V_r-E_l-J\tau_m)/\sigma \label{LIF_Fir_Rate_with_Noise2}\\
x_+ &= (V_t-E_l-J\tau_m)/\sigma \label{LIF_Fir_Rate_with_Noise3}
\end{align}
where $erfc$ is the complementary error function. Figure \ref{F-I_noise_dep} shows that both predictions are quite accurate in predicting the mean firing rate. 
\begin{figure}

\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/F-I_noise_dep.png}
\end{center}
\caption{Comparison of mean firing rate of a LIF neuron without noise (squares) and with noise (circles, $\mathrm{\sigma = \sqrt{5}\, mV}$), averaged over $\mathrm{50\, s}$. Red and blue curves are the corresponding analytic predictions, see equation \ref{LIF_Fir_Rate_no_Noise} and \ref{LIF_Fir_Rate_with_Noise1}.}
\label{F-I_noise_dep}
\end{figure}
However, one should note that the synaptic input within a spiking network carries an intrinsic randomness: since it is the sum of instantaneous increases in membrane voltage upon arriving spikes, one cannot simply describe it as a mean constant input $J$. Rather, if we assume that the arrival of spikes is approximately Poisson-distributed and the time constant of the membrane is small compared to the average interspike-interval, according to \cite{Roxin_Firing_Rate_Distribution}, one can describe the synaptic input from population $\mathrm{A}$ to population $\mathrm{B}$ as $J_{AB}(t) = \mu_{AB} (t) + \sigma_{AB} \cdot \zeta(t)$, where $\zeta(t)$ is zero-mean standard gaussian noise and
\begin{align}
\mu_{AB} &= <w_{AB}> \cdot <N_{in,AB}> \cdot <r_A> \label{Noise_Approx_1}\\
\sigma_{AB} &= \sqrt{\tau_m \cdot <w_{AB}>^2 \cdot <N_{in,AB}> \cdot <r_A>} \label{Noise_Approx_2} \, .
\end{align}
$\mathrm{<w_{AB}>,\, <N_{in,AB}>}$ and $\mathrm{<r_A>}$ denote the mean weight of a synapse connection population $\mathrm{A}$ with population $\mathrm{B}$, the mean number of incoming connections at a neuron in population $\mathrm{B}$ and the mean firing rate in population $\mathrm{A}$ (which makes it necessary to assume changes within the firing rate to be slow enough for a quasi-stationary description of the activity).
Since we can assume intrinsic and input noise to be uncorrelated, the total noise within a neuron of population $\mathrm{B}$ has a standard deviation of 
\begin{equation}
\sigma_{tot.} = \sqrt{\sigma_{intr.}^2 + \sum_A \sigma_{AB}^2} \; .
\label{Total_Noise_Neuron}
\end{equation}

Apart from the correct description of $\phi_{e/i}$, a third problem is the analytic description of the STP that was present for recurrent excitatory connections in our network. In \cite{Markram_STP}, the steady-state value of $\mathrm{x\cdot u}$ for an input of constant rate $\mathrm{r}$ is given by
\begin{align}
x_{st.}(r) &= \frac{1-\exp \left(-\frac{1}{r\cdot \tau _d}\right)}{1-\left(1-u_{st.}\left(r\right)\right)\cdot \exp \left(-\frac{1}{r\cdot \tau _d}\right)} \label{STP_steady1} \\
u_{st.}(r) &= \frac{U}{1-\left(1-U\right)\exp \left(-\frac{1}{r\cdot \tau _f}\right)} \; .\label{STP_steady2}
\end{align}
When trying to find the fixed point of equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, one thus needs to incorporate the factor $\mathrm{x_{st.}\cdot u_{st.}}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations results in a stable configuration depicted in \ref{Dyn_Rate_Approx}. For this simulation, we set the relevant parameters to their predefined values and, in addition, extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, being the mean excitatory and inhibitory firing rate between 100-750 s of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
%% fixed_point_net.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that were present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principal, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $\mathrm{(r_{0,e},r_{0,i})}$. For the sake of the initial question though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: Assuming validity of the described rate-model, the recurrent network as such (without homeostasis) is stable under the given choice of parameters. Furthermore, on an empirical basis, the occurrence of oscillations appear to be rather dependent on the choice of homeostasis, since apart from the homeostatic mechanism itself, all parameters of the simulation were kept unchanged upon the transition after 750 s. In consequence, the following section further analyzes the dynamics underlying the diffusive homeostatic feedback.

@@

Therefore, the resulting rate of NO synthesis can be decomposed into a sum of time shifted responses onto a single kernel of Calcium concentration as a result of a spike. for a spike at $t_{spike}=0$, the solution of \eqref{nNOS_dyn} can be calculated by

@@

The exact solution of this integral can be expressed in terms of the hyper-geometric function, making it rather impractical for any further analysis. Looking for further simplifications, we note that $\tau_{nNOS}$ is chosen ten-fold larger than $\tau_{Ca^{2+}}$. This discrepancy in decay times allows for the assumption that the impact of the Calcium kernel onto $nNOS$ is practically instantaneous. Consequently, $nNOS(t)$ becomes

@@

As Figure \ref{Fir_Rate_Dist_Sim_vs_Mat} shows, the actual shape of the distribution is well predicted by the solution of the linear system.

>>>>

@@

\begin{document}
\begin{titlepage}
\clearpage
\thispagestyle{empty}
\begin{center}
\LARGE{{\scshape Faculty 13 Physics}\\
  {\scshape Frankfurt Institute for Advanced Studies}\\
  \vspace{1cm}
  Master's Thesis\\}
  \vspace{1cm}
  \huge{Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network}\\
  \vspace{0.5cm}
  \LARGE{Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity}\\
  \vspace{2cm}
  \LARGE{Fabian Schubert\\
  Matriculation Number: 4282141}\\
  \vspace{2cm}
  Thesis Supervisor: Prof. Dr. Jochen Triesch\\
  \end{center}

\vfill

% Bottom of the page
\begin{center}
{\large \today}
\end{center}
\end{titlepage}

\clearpage
\thispagestyle{empty}

\begin{LARGE}
\noindent\textbf{Statutory Declaration}
\end{LARGE}
\vspace{.5cm}\\
I hereby declare that I authored this thesis independently and that I have
not used other than the declared sources/resources. I explicitly marked all material which has been quoted either literally or by content from the used sources. I further declare that neither the thesis, nor any section of it, was used as a study achievement for any other graded or ungraded exam.  
\vspace{2cm}\\
\rule{\textwidth}{0.4pt}

\clearpage
\thispagestyle{empty}

\tableofcontents
\newpage
\clearpage
\setcounter{page}{1}

\makeatother
\MSonehalfspacing

\section{Introduction}
Among other functional properties of the mammalian central nervous system, it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found up till now, and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though, experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons being more or less active than a certain set point of mean activity are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. Furthermore, one can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing "intrinsic" properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}. It is important to note that - depending on the specific mechanisms involved - both concepts of neural homeostasis can occur on a very wide range of different time scales, ranging from milliseconds to hours or days.

@@

Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity which has previously been suppressed by the rigid single-cell homeostasis. In the face of possible \emph{new} features within the network's structure, we will clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

Equation \eqref{NO_dyn} describes the influx of NO as a sum of scaled and spatially shifted Dirac functions. Apart from the question, whether this source term results in a well defined, finite solution at the neurons' positions (see section \ref{Matrix_Diff_Model_Section}), it can only be modeled to a certain degree of accuracy, depending on the resolution of the numeric grid. In practice, we approximated the point sources of NO as an insertions at individual grid cells at a rate of $nNOS^i(t) / h^2$, where the normalizing divisor $h^2$ ensured the desired total influx per neuron. This numeric implementation required two additional constraints: First, all random neuron positions were confined to integer multiples of $h$ in x- and y-direction. Second, to avoid redundancy and for physiological reasons, each grid cell could only hold one neuron at maximum. Both conditions combined led to an iterative generation of positions, where for each neuron a random number generator produced pairs of integers $(n_x,n_y)$, each within $[0,N]$, until an unoccupied pair of integers was found and occupied, moving on to the next neuron. Figure \ref{diff_test_plot} shows an example of the resulting NO density throughout a simulation. 
\begin{figure}[h!]
%% density_plot.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/density_plots/density_plot.png}
\end{center}
\caption{Example of NO-diffusion with 400 point sources of excitatory neurons.}
\label{diff_test_plot}
\end{figure}

@@

\subsection{Activity Analysis} \label{activ_analys}
As a first attempt, we set the time constant of threshold adaptation in the diffusive homeostasis to $\mathrm{2500\, ms}$, as given by \cite{Sweeney_Paper}. Switching between non-diffusive/diffusive homeostasis after 750 seconds, Figure \ref{full_sim_osci} shows the resulting dynamics of the population activity of the excitatory and inhibitory group, the average $NO$ concentration at the excitatory neurons' positions and the average firing threshold within the excitatory group (inhibitory threshold was fixed, see table \ref{LIF_neuron_params}). Both homeostatic mechanisms managed to keep the excitatory population activity in the desired range of 3 Hz. However, what might appear to be slightly faster and stronger random fluctuations in the upper three plots of Figure \ref{full_sim_osci}, were in fact regular oscillations across all three depicted variables. While the oscillation amplitudes undergo a rather unpredictable course, the frequency remains at a constant level of $\simeq 0.5\, Hz$. This feature is also illustrated by the fact that the power spectra depicted in Figure \ref{Power_Spec_without_Analysis} have a peak at the frequency that appeared to be dominant in Figure \ref{full_sim_osci}. The fact that the mean over all power spectra of excitatory thresholds differs from the power spectrum of the mean of these thresholds in its amplitude indicates that not all thresholds oscillate at the same phase. Still though, the overall shape of both spectra are equivalent. 
\begin{figure}
\begin{center}
%% oscillating_activity_plot.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl.png}
\end{center}
%\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_closeup.png}
\caption{A/A*: Mean of excitatory (blue) and inhibitory (red) neuronal firing rate. B/B*: NO concentration averaged over readouts at excitatory neurons' positions. C/C*: Average excitatory firing threshold. Non-diffusive homeostasis was used for $\mathrm{0-700\, s}$ , diffusive h. for $\mathrm{700-1500\, s}$. Target activity was $\mathrm{3\, Hz}$.}
\label{full_sim_osci}
\end{figure}
\begin{figure}
\begin{center}
%%% plot_pow_spec.py
\includegraphics[width=0.7\textwidth]{../../plots/power_spec/power_spectrum_sim_only.png}
\caption{Blue: Mean of power spectrum over all excitatory thresholds. Red: Power spectrum of mean over all excitatory thresholds.}
\label{Power_Spec_without_Analysis}
\end{center}
\end{figure}
Although one might argue that - for the sake of practical purposes - regular oscillations of the $NO$ concentration are not of much concern for the network as such, a periodically fluctuating firing rate is a qualitative feature requiring further inquiry. Rhythmic patterns of activity are a commonly observed phenomenon within the brain. Depending on the rate of oscillations, the location within the brain and the spatial scale over which they are observed, different physiological features are found to be responsible, leading to explanatory models at different levels of abstraction. On a microscopic, single-neuron level, a so-called slow after-hyper-polarization conductance explains the appearance of bursting behavior: A relatively short period of spiking activity is followed by a longer period of silence. This phenomenon can be modeled by a $Ca^{2+}$-dependent $K^{+}$-conductance. Upon the bursting period, $Ca^{2+}$ concentration rises, prohibiting continuous spiking due to the activation of $K^{+}$  current until the $Ca^{2+}$ concentration has returned to base-level \cite[p.~203-207]{Theor_Neur_Dayan}. However, due to the simplicity of our neuron model, this type of oscillating activity can be ruled out as an explanation.

On a larger, rate-based level, dynamical models of interconnected excitatory and inhibitory groups can exhibit sustained oscillations of activity \cite[p.~270]{Theor_Neur_Dayan}. As a simplest model, one can describe the dynamics of the excitatory and inhibitory populations as a two-dimensional dynamical system, representing the respective population rates $r_e$ and $r_i$:
\begin{align}
\tau_e \dot{r}_e  &= -r_e + \phi_e(r_e W_{ee} + r_i W_{ie}) \label{2d-exc_inh_rate_1} \\
\tau_i \dot{r}_i  &= -r_i + \phi_i(r_e W_{ei} + r_i W_{ii}) \label{2d-exc_inh_rate_2}
\end{align}

where $\tau_e$ and $\tau_i$ are time constants representing the rapidity of firing rate adaptation, $\phi_e$ and $\phi_i$ the respective transition functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the 2-dimensional phase plane. According to the Poincaré-Bendixson theorem, such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing "into" $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.
Coming back to the dynamics in question, one would need to show that the network has a locally unstable fixed point at $\mathrm{(r_e,r_i)\approx (3\, Hz, 13\, Hz)}$, see A* in Figure \ref{full_sim_osci}. Several aspects come into play, making it a hard task to fit the full spiking network model onto the simplified equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. First, the choice of $\tau_e$ and $\tau_i$ in \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} has significant influence onto the stability of the system. However, a straightforward equivalent does not exist in the used spiking neuron model. It has been shown that in the case of slowly varying input $\tau_{e/i}$ of equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} are equal to the membrane constant $\tau_m$ of the LIF-neuron used in the network \cite{Gerstner_Pop_Act}, see equation \ref{LIF_Dynamics}. A second problem is to choose a good estimate of the activation functions $\phi_{e/i}( \cdot )$. In the absence of noise in equation \ref{LIF_Dynamics}, the firing rate for a given constant total input $J$ is
\phi(J) = \begin{cases}\frac{1}{-\tau_m ln(\frac{V_t -E_l-J\tau_m}{V_r -E_l-J\tau_m})} & \mathrm{for}\; J\tau_m + E_l > V_t\\
0 & \mathrm{otherwise}
\end{cases}.
\label{LIF_Fir_Rate_no_Noise}
If noise $\sigma$ is present, according to \cite{Roxin_Firing_Rate_Distribution} one can calculate the \emph{mean} firing rate by
\phi(J) &= \left[ \sqrt{\pi}\tau_m \int_{x_-}^{x_+} dx e^{x^2} erfc(-x) \right]^{-1} \label{LIF_Fir_Rate_with_Noise1}\\
x_- &= (V_r-E_l-J\tau_m)/\sigma \label{LIF_Fir_Rate_with_Noise2}\\
x_+ &= (V_t-E_l-J\tau_m)/\sigma \label{LIF_Fir_Rate_with_Noise3}
where $erfc$ is the complementary error function. Figure \ref{F-I_noise_dep} shows that both predictions are quite accurate in predicting the mean firing rate. 
\includegraphics[width=0.7\textwidth]{../../plots/F-I_noise_dep.png}
\caption{Comparison of mean firing rate of a LIF neuron without noise (squares) and with noise (circles, $\mathrm{\sigma = \sqrt{5}\, mV}$), averaged over $\mathrm{50\, s}$. Red and blue curves are the corresponding analytic predictions, see equation \ref{LIF_Fir_Rate_no_Noise} and \ref{LIF_Fir_Rate_with_Noise1}.}
\label{F-I_noise_dep}
However, one should note that the synaptic input within a spiking network carries an intrinsic randomness: since it is the sum of instantaneous increases in membrane voltage upon arriving spikes, one cannot simply describe it as a mean constant input $J$. Rather, if we assume that the arrival of spikes is approximately Poisson-distributed and the time constant of the membrane is small compared to the average interspike-interval, according to \cite{Roxin_Firing_Rate_Distribution}, one can describe the synaptic input from population $\mathrm{A}$ to population $\mathrm{B}$ as $J_{AB}(t) = \mu_{AB} (t) + \sigma_{AB} \cdot \zeta(t)$, where $\zeta(t)$ is zero-mean standard gaussian noise and
\mu_{AB} &= <w_{AB}> \cdot <N_{in,AB}> \cdot <r_A> \label{Noise_Approx_1}\\
\sigma_{AB} &= \sqrt{\tau_m \cdot <w_{AB}>^2 \cdot <N_{in,AB}> \cdot <r_A>} \label{Noise_Approx_2} \, .
$\mathrm{<w_{AB}>,\, <N_{in,AB}>}$ and $\mathrm{<r_A>}$ denote the mean weight of a synapse connection population $\mathrm{A}$ with population $\mathrm{B}$, the mean number of incoming connections at a neuron in population $\mathrm{B}$ and the mean firing rate in population $\mathrm{A}$ (which makes it necessary to assume changes within the firing rate to be slow enough for a quasi-stationary description of the activity).
Since we can assume intrinsic and input noise to be uncorrelated, the total noise within a neuron of population $\mathrm{B}$ has a standard deviation of 
\sigma_{tot.} = \sqrt{\sigma_{intr.}^2 + \sum_A \sigma_{AB}^2} \; .
\label{Total_Noise_Neuron}
Apart from the correct description of $\phi_{e/i}$, a third problem is the analytic description of the STP that was present for recurrent excitatory connections in our network. In \cite{Markram_STP}, the steady-state value of $\mathrm{x\cdot u}$ for an input of constant rate $\mathrm{r}$ is given by
x_{st.}(r) &= \frac{1-\exp \left(-\frac{1}{r\cdot \tau _d}\right)}{1-\left(1-u_{st.}\left(r\right)\right)\cdot \exp \left(-\frac{1}{r\cdot \tau _d}\right)} \label{STP_steady1} \\
u_{st.}(r) &= \frac{U}{1-\left(1-U\right)\exp \left(-\frac{1}{r\cdot \tau _f}\right)} \; .\label{STP_steady2}
When trying to find the fixed point of equations \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, one thus needs to incorporate the factor $\mathrm{x_{st.}\cdot u_{st.}}$ by means of the previously described mean input and the contribution to the total variance of the membrane noise.

All these aspects taken together make it impossible to find an analytic expression for the fixed point of Equation \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}. Running a simulation of \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2}, including all aforementioned approximations results in a stable configuration depicted in \ref{Dyn_Rate_Approx}. For this simulation, we set the relevant parameters to their predefined values and, in addition, extracted the mean threshold of the excitatory population which was set by the non-diffusive homeostasis, see C/C* in Figure \ref{full_sim_osci}, yielding a threshold of $\mathrm{-56.963\, mV}$. The results are depicted in Figure \ref{Dyn_Rate_Approx}. The system settled to a fixed point of $\mathrm{(r_{0,e},r_{0,i}) = (2.992\, Hz, 7.144\, Hz)}$, which is quite close to $\mathrm{(3\, Hz, 6.768\, Hz)}$, being the mean excitatory and inhibitory firing rate between 100-750 s of the full simulation (i.e., during non-diffusive homeostasis).
\begin{figure}
%% fixed_point_net.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/approx_rates_EI.png}
\end{center}
\caption{Dynamics of equations \ref{2d-exc_inh_rate_1} (blue line) and \ref{2d-exc_inh_rate_2} (red line). Dotted lines mark the mean frequencies that were present in the full spiking network. Excitatory: 3 Hz, Inhibitory: 6,768 Hz. \ref{2d-exc_inh_rate_1} and \ref{2d-exc_inh_rate_2} converged to 2.992 Hz and 7.144 Hz respectively.}
\label{Dyn_Rate_Approx}
\end{figure}
In principal, one could now further quantify the stability of the found fixed point by calculating the Jacobian matrix at $\mathrm{(r_{0,e},r_{0,i})}$. For the sake of the initial question though, namely the source of oscillation, it is sufficient at this point to restrict oneself to a preliminary result: Assuming validity of the described rate-model, the recurrent network as such (without homeostasis) is stable under the given choice of parameters. Furthermore, on an empirical basis, the occurrence of oscillations appear to be rather dependent on the choice of homeostasis, since apart from the homeostatic mechanism itself, all parameters of the simulation were kept unchanged upon the transition after 750 s. Therefore, the following section further analyzes the dynamics underlying the diffusive homeostatic feedback.

@@

Therefore, the resulting rate of NO synthesis can be decomposed into a sum of time shifted responses onto a single kernel of calcium concentration as a result of a spike. for a spike at $t_{spike}=0$, the solution of \eqref{nNOS_dyn} can be calculated by

@@

The exact solution of this integral can be expressed in terms of the hyper-geometric function, making it rather impractical for any further analysis. Looking for further simplifications, we noted that $\tau_{nNOS}$ is ten-fold larger than $\tau_{Ca^{2+}}$. This discrepancy in decay times allows for the assumption that the impact of the calcium kernel onto $nNOS$ is practically instantaneous. Consequently, $nNOS(t)$ becomes

@@

As Figure \ref{Fir_Rate_Dist_Sim_vs_Mat} shows, the actual shape of the distribution was also well predicted by the solution of the linear system.
